Eye in a Disk: eyeIntegration human pan-eye and body transcriptome database version 1.00

### 2019-03-01 manuscript.Rmd
discussion updates

We plotted these three sets of genes in the heatmap view, with columns as age in days of fetal or organoid retina and rows are genes. We see that there are strong, consistent gene expression changes in these three groups of genes between fetal retina and the organoid samples (`r paste0(fig_cap(name = 'rgc_cadherin_hoxb', display = 'cite'))`). We also plot the differential expression differences between all organoids and all fetal retina samples; all genes across all three sets are significantly differentially expressed with a FDR corrected p value < 0.01.

@@

As human fetal tissue is difficult to obtain and thus not very amenable for chemical or genetic modification, it is crucial for iPSC- and organoid-based in vitro developmental models to be developed. Our merging of these datasets and analysis at the transcriptome level (as compared to cross-analyzing using a limited number of known marker genes) demonstrates that these two approaches successfully recapitulate fetal tissues to a first approximation at the whole transcriptome level. This result underscores the value of these approaches in building models for human retina development.

>>>>

We plotted these three sets of genes in the heatmap view, with columns as age in days of fetal or organoid retina. As above, rows are genes, split by the three groups of genes and clustered by how related the expression patterns are. We see that there are strong, consistent gene expression changes in these three groups of genes between fetal retina and the organoid samples (`r paste0(fig_cap(name = 'rgc_cadherin_hoxb', display = 'cite'))`). We also plot the differential expression differences between all organoids and all fetal retina samples; all genes across all three sets are significantly differentially expressed with a FDR corrected p value < 0.01.

@@

As human fetal tissue is difficult to obtain and thus not very amenable for chemical or genetic modification, it is crucial for organoid-based models to be developed. Our merging of these datasets and analysis at the transcriptome level (as compared to cross-analyzing using a limited number of known marker genes) demonstrates that these two approaches successfully recapitulate fetal tissues to a first approximation at the whole transcriptome level. However as organoids do yet develop to full function, it is important to look at how gene expression systems differ between organoid and fetal tissue. https://www.ncbi.nlm.nih.gov/pubmed/12514132

We use our large dataset to narrow in on some processes which to identify three core processes which differ. First we show that the HOXB family is overexpressed in the organoids. The homeobox family is well known to initiate polarity of the embryo during early development. As retinoic acid, which interacts directly to control expression of HOX genes, is used in the differentiation process from stem cell to organoids, it is likely that the HOXB expression is a by-product of the morphogen. While higher HOXB expression may be important to spur differentiation of the multiple cell types of the eye, the continued HOXB expression in the older organoids may inhibit the maturation of the organoid.

Next we demonstrate that many protocadherins are more highly expressed in the fetal tissue, relative to the organoids. As protocadherins mediate cell to cell connections and, in the develpoing mouse are shown to be important for spinal internneurons and RGC survival. We would predict that decreased protocadherin expression would reduce RGC activity, which is exactly the result we see. Many canonical RGC markers, while present in detectable levels in the organoids, are signficantly underexpressed in the organoids relative to the tissues. This result suggests that genetically modifying stem cells to increase protocadherin expression may result in higher RGC activity.

https://www.ncbi.nlm.nih.gov/pubmed/30266927

http://dev.biologists.org/content/135/24/4141?ijkey=2ce65c0994c48221620a90803e1861685b60935c&keytype2=tf_ipsecsha

http://jcs.biologists.org/content/128/8/1455

We built the Eye in a Disk dataset and the accompanying web app, eyeIntegration in the hopes that easily accessible gene expression across tissue space and time will be a
